Initial estuarine response to the nutrient-rich Piney Point discharge into Tampa Bay, Florida

### 2021-11-08 manu-draft.Rmd
edit to funding acknowledgements
EM and EP edits added

---
title: "Initial estuarine response to the nutrient dense Piney Point discharge into Tampa Bay, Florida"
output: 
  bookdown::word_document2:
    reference_docx: my_styles.docx
    number_sections: false
always_allow_html: true
bibliography: refs.bib
author: "Marcus W. Beck (mbeck@tbep.org), Andrew Altieri (andrew.altieri@essie.ufl.edu), Christine Angelini (christine.angelini@essie.ufl.edu), Maya C. Burke (mburke@tbep.org), Jing Chen (jchen15@usf.edu), Diana W. Chin (diwchin@gmail.com), Emma E. Dontis (edontis@co.pinellas.fl.us), Jayne Gardiner (jgardiner@ncf.edu), Jessica Lewis (jlewis@tbep.org), Yonggang Liu (yliu@usf.edu),  Cary Lopez (cary.lopez@myfwc.com), Miles Medina (miles.medina@ufl.edu), Elise Morrison (elise.morrison@essie.ufl.edu), Edward J. Phlips (phlips@ufl.edu), Gary E. Raulerson (graulerson@tbep.org), Sheila Scolaro (sscolar@tbep.org), Edward T. Sherwood (esherwood@tbep.org), David Tomasko (dave@sarasotabay.org), Bob Weisberg (weisberg@usf.edu), Nijole Wellendorf (nijole.wellendorf@floridadep.org), Joe Whalen (jwhalen@tbep.org)"
urlcolor: blue
csl: marine-pollution-bulletin.csl
link-citations: true
---

@@

Aging infrastructure and climate change stressors acting on wastewater containment facilities are global concerns for the environmental protection of surface waters.  From March 30th to April 9th, 2021, 814 million liters of mixed legacy phosphate mining wastewater and marine dredge water from the Piney Point facility were released into Tampa Bay (Florida, USA).  An estimated 186 metric tons of total nitrogen were exported to Lower Tampa Bay, exceeding typical annual external nitrogen load estimates in a matter of days.  An immediate phytoplankton response (non-harmful diatom) was observed in samples closest to the discharge site. Cyanobacteria blooms (*Dapis spp.*) peaked in June, followed by *Karenia brevis* blooms in Tampa Bay proper in July.  Reported fish kills tracked red tide concentrations, prompting local cleanup efforts to remove over 1600 metric tons of dead fish.  By comparing these observations to decades of baseline environmental monitoring data for the region and typical seasonal conditions experienced in recent years, we conclude that all of these biological responses observed after the discharge from Piney Point are abnormal relative to historic conditions in Tampa Bay.

@@

Long-term water quality monitoring data from Hillsborough and Manatee counties (accessible at https://wateratlas.usf.edu/) were used to establish baseline conditions for major areas of interest in Figure \@ref(fig:map)a to compare with the response monitoring data described above. These areas (Area 1: closest to Piney Point; Area 2: north of Piney Point; Area 3: south of Piney Point including northern Sarasota Bay) were identified based on anticipated impacts from expected plume patterns following the TBCOM simulations and other prominent bay boundaries relative to Piney Point (i.e., the main shipping channel in the bay, inflow boundaries, location of the Skyway Bridge at the mouth of Tampa Bay, and major bay segments used by TBEP for assessing programmatic goals). Observations at each monitoring station were averaged for each month across years from 2006 to 2020. This period represents a "recovery" stage for Tampa Bay where water quality conditions were much improved from historical conditions during a more eutrophic period and when seagrass areal coverage was trending towards and above a 1950s benchmark target of 15,378 hectares (38,000 acres) [@Greening14;@Sherwood17].  For each month, the mean values +/- 1 standard deviation for each parameter at each station were quantified and used as reference values relative to results at the closest monitoring station that was sampled in response to Piney Point.  This comparison was made to ensure that the response data were evaluated relative to stations that were spatially relevant (e.g., long-term conditions in Terra Ceia Bay are not the same as those in middle Tampa Bay) and seasonally-specific (e.g., historical conditions in April are not the same as historical conditions in July). In some cases, the nearest long-term station did not include data for every monitoring parameter at a response location and the next closest station was used as a reference (see https://shiny.tbep.org/piney-point/ for a map of the matches).

@@

A general summary of 2021 events in Tampa Bay following discharge rom Piney Point is shown in Figure \@ref(fig:timeline). After the discharge ceased on April 9th, an initial phytoplankton response was observed near Piney Point with concentrations peaking around mid-April.  Taxa from the Bacillariophyta phylum (diatoms) were dominant in April, with a maximum April Chl-a concentration of 50.5 ug/L, although median concentrations for each week in April were less than 10 ug/L. The initial diatom bloom did not persist past April and was partially dispersed by the passage of a strong weather front across the Tampa Bay region on April 10.  On April 20th, *K. brevis* was first observed near Anna Maria Sound at the southern edge of the mouth of Tampa Bay and reached bloom concentrations (>10,000 cells/L) by May 23rd, although bloom conditions were limited to lower Tampa Bay.  Also during May, *Dapis spp.* (filamentous cyanobacteria macroalgae) were observed at high abundances in Anna Maria Sound and near Port Manatee.  *Dapis* were observed in large floating mats on the surface and covering benthic and seagrass habitats at these locations.  By June 27th, fish kill reports attributed to red tide increased with *K. brevis* cell concentrations in lower and middle Tampa Bay.  The center of tropical storm Elsa passed to the west of Tampa Bay on July 5th, causing a shift in prevailing winds from the southeast.  This shift contributed to an increase in fish kill reports by moving dead fish closer to heavily populated areas of Tampa Bay, specifically near the cities of St. Petersburg and Tampa. Concentrations of *K. brevis* in middle and lower Tampa Bay peaked in early to mid July, with bloom conditions not observed in the bay after July.  Conditions were relatively stable in August and September compared to the months prior. A quantitative description of these events follows.

@@

Samples collected in the bay between April through September 2021 indicated that water quality conditions were unusual and outside of normal values expected for each month. A total of `r nwq` samples were collected and analyzed for chl-a, dissolved oxygen, TN, total phosphorus, total ammonia nitrogen, nitrate/nitrite, pH, salinity, Secchi depth, and temperature (Table \@ref(tab:wqsumtab)). The percentage of observations outside of the normal range (mean +/- 1 standard deviation from long-term data) varied by location and parameter.  For chl-a, `r chlabv[['1']]`\% of the observations were above the normal range for Area 1 located closest to the discharge point, whereas only `r chlabv[[2]]`\% and `r chlabv[[3]]`\% were above for Areas 2 (to the north) and 3 (to the south), respectively. Total nitrogen concentrations were above the normal range for `r tnabv[['1']]`\% of observations in Area 1, whereas concentrations were above for `r tnabv[['2']]`\% of observations in Area 2 and `r tnabv[['3']]`\% in Area 3.  Secchi observations were below the normal range for `r secblw[['1']]`\% of observations in Area 1 and for `r secblw[['2']]`\% and `r secblw[['3']]`\% of observations in Areas 2 and 3.  Notable differences were also observed for dissolved oxygen (e.g., `r dosabv[['1']]`\% were above in Area 1, `r dosabv[['2']]`\% in Area 2).  Physical parameters (salinity, temperature) and inorganic nitrogen (ammonia, nitrate/nitrite) were more often in normal range, although initial time series showed much higher concentrations for ammonia in April near Area 1.  Ammonia concentrations were observed in excess of 10 mg/L, about three orders of magnitude above baseline (see supplement, similar to the discharge measurements in Table \@ref(tab:stktab)). Inorganic nitrogen did not persist at high concentrations past April as it was likely rapidly utilized by phytoplankton (see below).  Spatial variation among the parameters showed that values were generally above the normal range (or below for Secchi depth) for many locations near Piney Point, Anna Maria Sound, and the northern mouth of Tampa Bay (Figure \@ref(fig:wqmap)).

@@

A total of `r ntrns` transects were sampled for macroalgae and seagrass from April through September, each visited on average `r avevis` times per month.  Macroalgae observed along the transects varied in coverage, with red macroalgae groups having the highest frequency occurrence of `r foall['Red']`\%.  Common taxa in the red group included *Gracilaria spp.* and *Acanthophora sp.*.  Green and cyanobacteria macroalgae were less common, with frequency occurrences of `r foall['Green']`\% and `r foall['Cyanobacteria']`\%.  Common taxa in the green group included *Ulva spp.* and *Caulerpa spp.*, whereas *Dapis spp.* was the only cyanobacteria macroalgae observed.  Brown macroalgae (primarily in the genus *Feldmannia*) were only observed at `r foall['Brown']`\% frequency occurrence. For seagrasses, turtle grass (*T. testudinum*) was the dominant species with frequency occurrence of `r foall['Thalassia testudinum']`\% across all locations and sample dates.  Manatee grass (*S. filiforme*) and shoal grass (*H. wrightii*) had similar coverage across all transects, with frequency occurrences of `r foall['Syringodium filiforme']`\% and `r foall['Halodule wrightii']`\%, respectively. The frequency occurrences of seagrasses near Piney Point were similar to the long-term record of seagrass transect data available for Tampa Bay [@Sherwood17], with turtle grass being the dominant species in more euhaline waters closer to the Gulf. There is no historical macroalgae record for Tampa Bay that is comparable to the resolution of the 2021 samples.

A typical pattern for macroalgae and seagrass observed at many of the transects is shown in Figure \@ref(fig:trnex). Transect S3T6 is located less than one kilometer to the north of Port Manatee. Macroalgal abundances changed over the course of sampling similar to the remainder of transects sampled during the study.  Red macroalgae were present in high abundances from April to May.  *Dapis spp.* was first observed on May 24th and was present at all of the sample locations along this transect on June 4th and 15th.  *Dapis spp.* persisted through June and July, but not was observed after July 20th.  Green macroalgae taxa were first observed in July, although at generally low abundances. Red macroalgae were the dominant taxa by the end of September. Overall abundance of seagrass did not change from April 22nd through September. The site is dominated by manatee grass that was observed at nearly all of the sample points along the transect at varying coverages.

Monthly summaries in frequency occurrence by area (Figure \@ref(fig:trnfrq)) provided an indication of macroalgae and seagrass trends across all transects.  No transects were sampled in Area 2 to the north of Piney Point and no transects were sampled past September in Area 1.  Macroalgal dominance varied across the months, similar to the example in Figure \@ref(fig:trnex).  Red macroalgae was the dominant group across all months and areas, with the highest frequency occurrences observed in April (`r fomoarea1apr['Red']`\% in Area 1, `r fomoarea3apr['Red']`\% in Area 3). Cyanobacteria frequency occurrence peaked in June, with greater coverage in Area 3 (`r fomoarea3jun['Cyanobacteria']`\%) compared to Area 1 (`r fomoarea1jun['Cyanobacteria']`\%).  Green macroalgae had the second lowest frequency occurrence, although it increased slightly by the end of the study period (`r fomoarea1sep['Green']`\% in September in Area 1, `r fomoarea3oct['Green']`\% in October in Area 3).  Brown macroalgae was only observed at one transect. For seagrass, both areas had generally stable total frequency occurrence. Turtle grass (*T. testudinum*) occurred in higher frequency occurrence in both areas (`r fomoarea1ave['Thalassia testudinum']`\% overall in Area 1, `r fomoarea3ave['Thalassia testudinum']`\% overall Area 3), compared to shoal grass (*H. wrightii*, `r fomoarea1ave['Halodule wrightii']`\% Area 1, `r fomoarea3ave['Halodule wrightii']`\% Area 3) and manatee grass (*S. filiforme*, `r fomoarea1ave['Syringodium filiforme']`\% Area 1, `r fomoarea3ave['Syringodium filiforme']`\% Area 3).  Slight changes in frequency occurrence in Area 3 were observed for all species starting in July, with a slight reduction in frequency occurrence of turtle grass and an increase in shoal grass and manatee grass. Statistical analyses with multiple comparison tests confirmed the general trends described above (see supplement).

@@

The increase in *K. brevis* from April to July to concentrations exceeding 10,000 cells/L (minimum bloom concentration based on FWC breakpoints) was an anomaly in 2021 that is not regularly observed in Tampa Bay.  The historical record from 1953 to present in Figure \@ref(fig:redtide)a shows cell concentrations sampled in Tampa Bay between April and September, with only a handful of years having cell concentrations greater than 10,000 cells/L, notably 1963, 1971, 2005, 2018, and 2021. Median cell concentrations for most years were well below 1,000 cells/L. The highest concentration of 20 million cells/L was observed in 1971, whereas the second highest concentration of 17.6 million cells/L was observed in 2021. Bloom concentrations in 2021 were not observed until the week of May 23, with concentrations peaking by the week of July 4th, after which concentrations declined (Figure \@ref(fig:redtide)b).  Cumulative rainfall and associated inflow from the main rivers entering Tampa Bay were below historical values (1995 - 2020) in the months prior to the highest bloom concentrations (Figure \@ref(fig:redtide)c, d). This likely contributed to elevated salinity values in lower and middle Tampa Bay that created conditions favorable for *K. brevis* growth, in addition to the elevated nutrient concentrations from the Piney Point discharge.

@@

The observed conditions in Tampa Bay in 2021 following the discharge from Piney Point provide multiples lines of evidence for an adverse environmental response to a large pulse of inorganic nitrogen into the system.  Collectively, these observations show that conditions in 2021 were anomalous when compared to long-term monitoring data for Tampa Bay.  These anomalous events included 1) a large diatom bloom in April in the vicinity of the discharge at Port Manatee, 2) high abundance of Cyanobacteria in Anna Maria Sound and near Port Manatee, 3) bloom concentrations of red tide in lower and middle Tampa Bay from June through July, and 4) high incidence of fish kill reports prompting local governments to remove over 1600 metric tons of dead fish from shoreline areas.  The observed water quality conditions throughout this period, particularly for TN, chl-a, and Secchi depth, were outside of normal seasonal ranges for many of the observations (Figures \@ref(fig:wqmap), Table \@ref(tab:wqsumtab)).  These conditions were widespread in lower and middle Tampa Bay.

@@

Several of the results of the water quality responses are consistent with observations of nutrient loading in other shallow Gulf Coast estuaries [@Doering06;@Caffrey14;@Greening14].  Regarding the nutrient management paradigm for Tampa Bay, the relationship between nutrients, chl-a, and water transparency followed expectations of reduced water quality with increased nutrient loads. Temporally, these changes were observed at different times and for different species of phytoplankton.  The initial increase in chl-a was first associated with a diatom bloom in April and then by *K. brevis* in June and July.  Peaks in dissolved oxygen saturation were also observed as an indicator of elevated phytoplankton production [@Kemp80], particularly in July with the peak *K. brevis* bloom.  Of note is that inorganic species of nitrogen, mainly ammonia, were only present at high concentrations in early April.  Concerns of the negative impacts of nutrients focused primarily on the high concentrations of ammonia in the discharge (Table \@ref(tab:stktab)), which can be utilized rapidly by many phytoplankton taxa [@Bates76;@Domingues11].  Low concentrations of ammonia after April may be explained by quick uptake by the initial diatom bloom, where TN that included organic sources were at high concentrations through April and again peaking in July.  Variation in observed concentrations of nutrients is complex given that high concentrations may suggest availability to support phytoplankton growth, whereas low concentrations may imply cycling of available nitrogen in organic forms already utilized by different taxa, including macroalgae [@Valiela97;@Cohen06].

Macroalgae trends across the study period were much more dramatic than the subtle differences observed in the seagrass community.  This was expected given both the documented changes from past releases from Piney Point [@Switzer11] and the more rapid response of macroalgae to changing water quality conditions relative to seagrasses [@Valiela97]. In Tampa Bay, red macroalgae groups (e.g., *Gracilaria spp.*, *Acanthophora sp.*) are more common than green macroalgae (e.g., *Ulva spp.*, *Caulerpa spp.*) and occur earlier in the growing season.  The dominance of the red groups early in the summer followed by an increase in the green alga *Ulva spp.* may reflect a natural phenology in Tampa Bay. The most notable change in the macroalgae community was a high abundance of cyanobacteria (i.e., *Dapis spp.*) in May and June.  Unusually high abundances of *Dapis spp.* were observed in Anna Maria Sound near the mouth of Tampa Bay and near Port Manatee at the discharge site. Long-term monitoring data describing normal seasonal variation in macroalgae are unavailable and we cannot distinguish between seasonal changes and those as a potential response to the Piney Point discharge. Cyanobacteria has been observed during routine annual transect monitoring in Tampa Bay and it has previously been documented in public reports to the Florida Department of Environmental Protection. However, these communities can respond rapidly to external nutrient inputs [@Albert05;@Ahern07], often exhibiting lagged responses with characteristic growth/decay periods similar to observations herein [@Estrella13], and it is not unreasonable to expect these trends to be related to nutrients from Piney Point.

>>>>

---
title: "Initial estuarine response to the nutrient-rich Piney Point discharge into Tampa Bay, Florida"
output: 
  bookdown::word_document2:
    reference_docx: my_styles.docx
    number_sections: false
always_allow_html: true
bibliography: refs.bib
author: "Marcus W. Beck (mbeck@tbep.org), Andrew Altieri (andrew.altieri@essie.ufl.edu), Christine Angelini (christine.angelini@essie.ufl.edu), Maya C. Burke (mburke@tbep.org), Jing Chen (jchen15@usf.edu), Diana W. Chin (diwchin@gmail.com), Emma E. Dontis (edontis@co.pinellas.fl.us), Jayne Gardiner (jgardiner@ncf.edu), Jessica Lewis (jlewis@tbep.org), Yonggang Liu (yliu@usf.edu),  Cary Lopez (cary.lopez@myfwc.com), Miles Medina (miles.medina@ufl.edu), Elise Morrison (elise.morrison@essie.ufl.edu), Edward J. Phlips (phlips@ufl.edu), Gary E. Raulerson (graulerson@tbep.org), Sheila Scolaro (sscolar@tbep.org), Edward T. Sherwood (esherwood@tbep.org), David Tomasko (dave@sarasotabay.org), Bob Weisberg (weisberg@usf.edu), Nijole Wellendorf (nijole.wellendorf@floridadep.org), Joe Whalen (jwhalen@tbep.org)"
urlcolor: blue
csl: marine-pollution-bulletin.csl
link-citations: true
---

@@

Aging infrastructure and climate change stressors acting on wastewater containment facilities are global concerns for the environmental protection of surface waters.  From March 30th to April 9th, 2021, 814 million liters of mixed legacy phosphate mining wastewater and marine dredge water from the Piney Point facility were released into Tampa Bay (Florida, USA).  An estimated 186 metric tons of total nitrogen were exported to Lower Tampa Bay, exceeding typical annual external nitrogen load estimates in a matter of days.  In the period immediately after the discharge, elevated levels of phytoplankton (non-harmful diatoms) were observed in the nearshore regions of the lower Bay. Filamentous cyanobacteria blooms (*Dapis spp.*) peaked in June, followed by *Karenia brevis* blooms that peaked in July.  Reported fish kills tracked red tide concentrations, prompting local cleanup efforts to remove over 1600 metric tons of dead fish.  By comparing these observations to decades of baseline environmental monitoring data for the region and typical seasonal conditions experienced in recent years, we conclude that all of these biological responses observed after the discharge from Piney Point are abnormal relative to historic conditions in Tampa Bay.

@@

Long-term water quality monitoring data from Hillsborough and Manatee counties (accessible at https://wateratlas.usf.edu/) were used to establish baseline conditions for major areas of interest in Figure \@ref(fig:map)a to compare with the response monitoring data described above. These areas (Area 1: closest to Piney Point; Area 2: north of Piney Point; Area 3: south of Piney Point including northern Sarasota Bay) were identified based on anticipated impacts from expected plume patterns following the TBCOM simulations and other prominent bay boundaries relative to Piney Point (i.e., the main shipping channel in the bay, inflow boundaries, location of the Skyway Bridge at the mouth of Tampa Bay, and major bay segments used by TBEP for assessing programmatic goals). Observations at each monitoring station were averaged for each month across years from 2006 to 2020. This period represents a "recovery" stage for Tampa Bay where water quality conditions were much improved from historical conditions during a more eutrophic period and when seagrass areal coverage was trending towards and above a 1950s benchmark target of 15,378 hectares (38,000 acres) [@Greening14;@Sherwood17].  For each month, the mean values +/- 1 standard deviation for each parameter at each station were quantified and used as reference values relative to results at the closest monitoring station that was sampled in response to Piney Point.  This comparison was made to ensure that the response data were evaluated relative to stations that were spatially relevant (e.g., long-term conditions in Terra Ceia Bay are not the same as those in middle Tampa Bay) and seasonally-specific (e.g., historical conditions in April are not the same as historical conditions in July). In some cases, the nearest long-term station did not include data for every monitoring parameter at a response location and the next closest station was used as a reference.  The average distance from a monitoring location in 2021 to the long-term sites was 1.6 km (see https://shiny.tbep.org/piney-point/ for a map of the matches).

@@

A general summary of 2021 events in Tampa Bay following discharge from Piney Point is shown in Figure \@ref(fig:timeline). After the discharge ceased on April 9th, a peak in median chl-a concentration was observed near Piney Point (Area 1, Figure \@ref(fig:map)a) mid-April, with peak individual sample values in excess of 50 ug/L. Median concentrations for each week in April were less than 10 ug/L.  The discharge phytoplankton biomass was comprised of over 99% of a spherical nanoplanktonic chlorophyte, with density of 337 x 10$^6$ cells/L. The phytoplankton communities near the discharge area in April were generally dominated by diatoms (Bacillariophyta). The initial diatom bloom did not persist past April.  On April 20th, the toxic red tide species *Karenia brevis* was first observed near Anna Maria Sound at the southern edge of the mouth of Tampa Bay.  By May/June, bloom levels of *K. brevis* were observed, with peak concentrations in excess of 1 x 10$^6$ cells/L in lower Tampa Bay.  Also during May/June, high abundances of benthic filamentous cyanobacteria (*Dapis spp.*) were observed in Anna Maria Sound (Area 3) and near Port Manatee (Area 1).  High levels of cyanobacteria coverage on benthic and seagrass habitats were observed, in addition to large floating mats on the surface. By June 27th, fish kill reports attributed to red tide increased with *K. brevis* cell concentrations in lower and middle Tampa Bay.  The center of tropical storm Elsa passed to the west of Tampa Bay on July 5th, causing a shift in prevailing winds from the southeast.  This shift contributed to an increase in fish kill reports by moving dead fish closer to heavily populated areas of Tampa Bay, specifically near the cities of St. Petersburg and Tampa. Concentrations of *K. brevis* in middle and lower Tampa Bay peaked in early to mid July, with bloom conditions not observed in the bay after July.  Conditions were relatively stable in August and September compared to months prior. A quantitative description of these events follows.

@@

Samples collected in the bay between April through September 2021 indicated that water quality conditions were unusual and outside of normal values expected for each month. A total of `r nwq` samples were collected and analyzed for chl-a, dissolved oxygen, TN, total phosphorus, total ammonia nitrogen, nitrate/nitrite, pH, salinity, Secchi depth, and temperature (Table \@ref(tab:wqsumtab)). The percentage of observations outside of the normal range (mean +/- 1 standard deviation from long-term data) varied by location and parameter.  For chl-a, `r chlabv[['1']]`\% of the observations were above the normal range for Area 1 located closest to the discharge point, whereas only `r chlabv[[2]]`\% and `r chlabv[[3]]`\% were above for Areas 2 (to the north) and 3 (to the south), respectively. Total nitrogen concentrations were above the normal range for `r tnabv[['1']]`\% of observations in Area 1, whereas concentrations were above for `r tnabv[['2']]`\% of observations in Area 2 and `r tnabv[['3']]`\% in Area 3.  Secchi observations were below the normal range for `r secblw[['1']]`\% of observations in Area 1 and for `r secblw[['2']]`\% and `r secblw[['3']]`\% of observations in Areas 2 and 3.  Notable differences were also observed for dissolved oxygen (e.g., `r dosabv[['1']]`\% were above in Area 1, `r dosabv[['2']]`\% in Area 2).  Physical parameters (salinity, temperature) and inorganic nitrogen (ammonia, nitrate/nitrite) were more often in normal range, although initial time series showed much higher concentrations for ammonia in April near Area 1.  Ammonia concentrations were observed in excess of 10 mg/L, about three orders of magnitude above baseline (see supplement, similar to the discharge measurements in Table \@ref(tab:stktab)). Inorganic nitrogen did not persist at high concentrations past April as it was likely rapidly utilized by phytoplankton (see below).  Spatial variation among the parameters showed that values were generally above the normal range (or below for Secchi depth) for many locations near Piney Point (Area 1), Anna Maria Sound (Area 3), and the northern mouth of Tampa Bay (Area 3, Figure \@ref(fig:wqmap)).

@@

A total of `r ntrns` transects were sampled for macroalgae and seagrass from April through September, each visited on average `r avevis` times per month.  Macroalgae observed along the transects varied in coverage, with red macroalgae groups having the highest frequency occurrence of `r foall['Red']`\%.  Common taxa in the red group included genera *Gracilaria* and *Acanthophora*.  Green and cyanobacteria macroalgae were less common, with frequency occurrences of `r foall['Green']`\% and `r foall['Cyanobacteria']`\%.  Common taxa in the green group included genera *Ulva* and *Caulerpa*, whereas cyanobacteria biomass was dominated by the benthic filamentous genera *Dapis*.  Brown macroalgae (primarily in the genus *Feldmannia*) were only observed at `r foall['Brown']`\% frequency occurrence. For seagrasses, turtle grass (*T. testudinum*) was the dominant species with frequency occurrence of `r foall['Thalassia testudinum']`\% across all locations and sample dates.  Manatee grass (*S. filiforme*) and shoal grass (*H. wrightii*) had similar coverage across all transects, with frequency occurrences of `r foall['Syringodium filiforme']`\% and `r foall['Halodule wrightii']`\%, respectively. The frequency occurrences of seagrasses near Piney Point were similar to the long-term record of seagrass transect data available for Tampa Bay [@Sherwood17], with turtle grass being the dominant species in more euhaline waters closer to the Gulf. There is no historical macroalgae record for Tampa Bay that is comparable to the resolution of the 2021 samples.

A typical pattern for macroalgae and seagrass observed at many of the transects is shown in Figure \@ref(fig:trnex). Transect S3T6 is located less than one kilometer to the north of Port Manatee. Macroalgal abundances changed over the course of sampling similar to the remainder of transects sampled during the study.  Red macroalgae were present in high abundances from April to May.  Benthic filamentous macroalgal cyanobacteria (*Dapis spp.*) was first observed on May 24th and was present at all of the sample locations along this transect on June 4th and 15th.  Filamentous cyanobacteria persisted through June and July, but was not observed in abundance after July 20th.  Green macroalgae taxa were first observed in July, although at generally low abundances. Red macroalgae were the dominant taxa by the end of September. Overall abundance of seagrass did not change from April 22nd through September. The site is dominated by manatee grass that was observed at nearly all of the sample points along the transect at varying coverages.

Monthly summaries in frequency occurrence by area (Figure \@ref(fig:trnfrq)) provided an indication of macroalgae and seagrass trends across all transects.  No transects were sampled in Area 2 to the north of Piney Point and no transects were sampled past September in Area 1 given dispersal patterns of the discharge from the TBCOM simulations.  Macroalgal dominance varied across the months, similar to the example in Figure \@ref(fig:trnex).  Red macroalgae was the dominant group across all months and areas, with the highest frequency occurrences observed in April (`r fomoarea1apr['Red']`\% in Area 1, `r fomoarea3apr['Red']`\% in Area 3). Cyanobacteria frequency occurrence peaked in June, with greater coverage in Area 3 (`r fomoarea3jun['Cyanobacteria']`\%) compared to Area 1 (`r fomoarea1jun['Cyanobacteria']`\%).  Green macroalgae had the second lowest frequency occurrence, although it increased slightly by the end of the study period (`r fomoarea1sep['Green']`\% in September in Area 1, `r fomoarea3oct['Green']`\% in October in Area 3).  Brown macroalgae was only observed at one transect. For seagrass, both areas had generally stable total frequency occurrence. Turtle grass (*T. testudinum*) occurred in higher frequency occurrence in both areas (`r fomoarea1ave['Thalassia testudinum']`\% overall in Area 1, `r fomoarea3ave['Thalassia testudinum']`\% overall Area 3), compared to shoal grass (*H. wrightii*, `r fomoarea1ave['Halodule wrightii']`\% Area 1, `r fomoarea3ave['Halodule wrightii']`\% Area 3) and manatee grass (*S. filiforme*, `r fomoarea1ave['Syringodium filiforme']`\% Area 1, `r fomoarea3ave['Syringodium filiforme']`\% Area 3).  Slight changes in frequency occurrence in Area 3 were observed for all species starting in July, with a slight reduction in frequency occurrence of turtle grass and an increase in shoal grass and manatee grass. Statistical analyses with multiple comparison tests confirmed the general trends described above (see supplement).

@@

Concentrations of the red tide species *K. brevis* in 2021 were not observed until the week of May 23, with concentrations peaking by the week of July 4th, after which concentrations declined (Figure \@ref(fig:redtide)b). 
The increase in *K. brevis* from April to July to concentrations exceeding 10,000 cells/L (minimum bloom concentration based on FWC breakpoints) was an anomaly in 2021 that is not regularly observed in Tampa Bay.  The historical record from 1953 to present in Figure \@ref(fig:redtide)a shows cell concentrations sampled in Tampa Bay between April and September, with only a few years having cell concentrations greater than 10,000 cells/L, notably 1963, 1971, 2005, 2018, and 2021. Median cell concentrations for most years were well below 1,000 cells/L. The highest concentration of 20 million cells/L was observed in 1971, whereas the second highest concentration of 17.6 million cells/L was observed in 2021.  Cumulative rainfall and associated inflow from the main rivers entering Tampa Bay were below historical values (1995 - 2020) in the months prior to the highest bloom concentrations (Figure \@ref(fig:redtide)c, d). This likely contributed to elevated salinity values in lower and middle Tampa Bay that created conditions favorable for *K. brevis* growth, in addition to the elevated nutrient concentrations from the Piney Point discharge.

@@

The observed conditions in Tampa Bay in 2021 following the discharge from Piney Point provide multiples lines of evidence for an adverse environmental response to a large pulse of inorganic nitrogen into the system.  Collectively, these observations show that conditions in 2021 were anomalous when compared to long-term monitoring data for Tampa Bay.  These anomalous events included 1) a large diatom bloom in April in the vicinity of the discharge at Port Manatee, 2) high abundance of Cyanobacteria in Anna Maria Sound and near Port Manatee, 3) bloom concentrations of red tide in lower and middle Tampa Bay from June through July, and 4) high incidence of fish kill reports prompting local governments to remove over 1600 metric tons of dead fish from shoreline areas.  The water quality conditions observed during the study period, particularly for TN, chl-a, and Secchi depth, were outside of normal seasonal ranges for many of the observations (Figures \@ref(fig:wqmap), Table \@ref(tab:wqsumtab)).  These conditions were widespread in lower and middle Tampa Bay.

@@

Several of the results of the water quality responses are consistent with observations of nutrient loading in other shallow Gulf Coast estuaries [@Doering06;@Caffrey14;@Greening14].  Regarding the nutrient management paradigm for Tampa Bay, the relationship between nutrients, chl-a, and water transparency followed expectations of reduced water quality with increased nutrient loads. Temporally, these changes were observed at different times and for different species of phytoplankton.  The initial increase in chl-a was first associated with a diatom bloom in April.  In June, a bloom of the red tide species *K. brevis* was introduced to Tampa Bay from the Gulf of Mexico, and persisted at high densities in the Bay through July. Peaks in dissolved oxygen saturation were also observed as an indicator of elevated phytoplankton production [@Kemp80], particularly in July with the peak *K. brevis* bloom.  Of note is that inorganic species of nitrogen, mainly ammonia, were only present at high concentrations in early April.  Concerns of the negative impacts of nutrients focused primarily on the high concentrations of ammonia in the discharge (Table \@ref(tab:stktab)), which can be utilized rapidly by many phytoplankton taxa [@Bates76;@Domingues11].  Low concentrations of ammonia after April may be explained by quick uptake by the initial diatom bloom, where TN that included organic sources were at high concentrations through April and again peaking in July.  Variation in observed concentrations of nutrients is complex given that high concentrations may suggest availability to support phytoplankton growth, whereas low concentrations may imply cycling of available nitrogen in organic forms already utilized by different taxa, including macroalgae [@Valiela97;@Cohen06].

There were also concerns that such releases may have contributed to the persistence and intensity of the red tide event that spread through a wide region of lower and middle Tampa Bay. Fisheries resources were impacted by *K. brevis* blooms in June and July (Figure \@ref(fig:redtide)).  Quantitative data on changes in nekton abundance and diversity in Tampa Bay in 2021 are forthcoming. Routine sampling of fisheries occurs monthly in Tampa Bay and a long-term record back to 1998 provides detailed information for the major bay segments. Results from the Tampa Bay Nekton Index showed a decline in fisheries resources following a significant red tide event that persisted for several months in lower Tampa Bay in 2005 [@Flaherty11;@Schrandt21]. Given the observed *K. brevis* concentrations in 2021 and the magnitude of fish kills, mandates for catch and release for popular sportfishes (*Sciaenops ocellatus*, *Cynoscion nebulosus*, and *Centropomus undecimalis*) have been extended through the fall.  For past Piney Point events, @Switzer11 evaluated nekton communities in Bishop Harbor from November 2003 to October 2004 following discharge to the subembayment.  Fish community structure and species composition did not differ compared to a pre-impact period, although bloom concentrations of HAB species (*Prorocentrum minimum*, *Heterosigma akashiwo*), including *K. brevis* and diatoms, were observed in Bishop Harbor during this time [@Garrett11].  Prior blooms were more localized than the 2021 bloom, potentially allowing fishes to migrate from harmful conditions.

For seagrasses, both major bloom events in 2021 produced unfavourable water quality conditions, although changes in frequency occurrence of seagrasses were minimal over the study. The long-term effects of the Piney Point discharge on the seagrass community remains uncertain. From 2018 to 2020, seagrass coverage declined by 16\% in Tampa Bay, with similar losses observed in Sarasota Bay (18\%), Lemon Bay (12\%), and Charlotte Harbor (23\%) to the south (Southwest Florida Water Management District, unpublished results).  These broader trends suggest regional drivers are affecting seagrass communities [e.g., variation in precipitation, @Tomasko20], yet local issues specific to individual bays also pose challenges to managing water quality and subtidal habitats. Recent seagrass losses in Sarasota Bay may be linked to decreased light availability from a persistent *K. brevis* bloom in 2018. Although the 2021 red tide in Tampa Bay was short-lived, potential long-term effects on seagrasses remain a concern. Ecosystem shifts from seagrass to macroalgae dominated communities are also a concern, both in 2021 and as observed at some locations in recent years from the annual transect monitoring results for Tampa Bay.  In particular, increasing abundance in recent years of the green algae *Caulerpa sp.* has been observed at long-term transects that were previously dominated by seagrass.  These changes may be indicative of broader ecosystem shifts concurrent with alteration of nutrient loads or system resilience at the expense of seagrass communities [@Lloret05;@Stafford06]. Acute stressors from short-term events, such as discharge from Piney Point, create additional and often preventable challenges in the context of broader issues affecting ecosystem health.
